Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging

### 2018-02-24 HRUIonGPU/1st submit/jrnl_article.tex
Adding HRUIonGPU article

For a three dimensional beamspace we show that processing rates supporting real-time cardiac ultrasound imaging are possible, meaning that images can be processed faster than the image acquisition rate for a wide range of parameters. Image quality is investigated in an \textit{in-vivo} cardiac dataset. These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with improved lateral resolution both in element-space and beamspace.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway.

@@

% denne må utelates i IUS-artikkelen
In recent work we have applied a GPU-based Capon beamformer for both active sonar imaging \cite{Buskenes, Buskenes2013} and cardiac ultrasound imaging \cite{asen2012}. Sonar has a less strict real-time requirement than medical ultrasound imaging, which means that most array configurations can be handled in real time using the GPU. In \cite{asen2012} and in this paper we show that Capon beamforming applied in medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates that not even a cluster of high-end GPUs can provide. Since the execution time of Capon beamforming is cubic with respect to the number of array elements, we need methods that reduce the computational complexity for large arrays. Various approaches to low-complexity Capon beamforming for medical ultrasound imaging have been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}. In this paper we extend the work done in \cite{asen2012} and \cite{Buskenes2013} by utilizing a beamspace transformation. The beamspace transform has been proposed used in ultrasound imaging by Nilsen and Hafizovic \cite{Nilsen2009} in order to maintain high frame rates as the number of elements increases while maintaining almost the same lateral resolution and sidelobe reduction capabilities of element-space Capon when applied in narrow-beam systems.
%The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging. 
%(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)

In the next section we present background information on Capon beamforming in both element-space and beamspace. We also give an introduction to the GPU computing model. In Section \ref{sec:meth} we give a summary of our parallel implementation of the element-space Capon beamformer, before the GPU implementation of the beamspace Capon beamformer is presented in Section \ref{sec:bs}. Benchmark results and resulting images for both implementations are presented in Section \ref{sec:bench} and \ref{sec:images}, together with a discussion of trading image quality for speed in Section \ref{sec:trade}. Finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.

@@

The conventional form of array beamforming in medical ultrasound imaging is delay-and-sum (DAS). Here, the signal recorded at element number $m \in [0,M-1]$ at time $n$ is delayed with an appropriate delay $\Delta_m[n]$ to focus and steer the ultrasound beam. This delayed data is here denoted $x_m[n]$. The beamformer output $z[n]$ is then the sum of all $M$ elements weighted with a factor $w_m$,
\begin{align}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n] = \vec{w}^H\vec{x}[n], \label{eq:z}
\end{align}
where $(\cdot)^H$ is the Hermitian transpose. The weight vector, or window, $\vec{w}$ is usually real in DAS beamforming, and is applied to trade resolution for a lower side lobe level. Although different windows can be used in different areas of the images, $\vec{w}$ is usually predefined.

As shown in Fig.\ \ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.   The Capon beamformer produces weights that minimize the beamformer output power while maintaining unity gain in the focus direction. The result is that interference impinging from other directions is suppressed \cite{Synnevag2007}. As with coarse-fine beamformers \cite{Thomenius}, the complex weights cause a micro-steering which is minute compared to the initial steering and focusing step, therefore the beamformer also works well for broadband ultrasound signals.

@@

\begin{figure}
\centerline{\includegraphics[width=3in]{gfx/beamforming_mv.eps}}
\caption{Capon beamforming. The impinging signal is demodulated, and aligned in phase by applying steering and focusing delays $\Delta_m[n]$. Adaptive weights, $\vec{w}[n]$, are then calculated based on in-phase data, and finally the output is formed by a Capon-weighted coherent sum.}
\label{fig:mvbf}
\end{figure}

@@

For an imaging system using narrow transmit and receive beams, like cardiac ultrasound imaging, the received signal will be concentrated in a narrow band in wavenumber space. Hence, a small percentage of the beams will contain almost all the received energy. Nilsen and Hafizovic concluded that as little as three beams around (and including) the transmit direction were adequate to produce results comparable to applying Capon beamforming in element space. Hence, for equally pitched arrays the inversion steps is found to be constant with respect to the number of array elements.

@@

In CUDA programming a kernel function is executed over a large set of threads, where the kernel function describes the work to be done in each thread. As depicted in Fig.\ \ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block (e.g. $32\times32$). Threads inside a block support low-cost synchronization if needed. From a CUDA perspective the GPU consist of one or more streaming multi-processors (SM), where each SM is capable of executing multiple arithmetic operations per clock cycle.
%32 or more concurrent multiplications or additions, or both, depending on the architecture. A group of blocks is scheduled to each SM until all blocks have been processed.

@@

\section{Parallel ES-Capon}\label{sec:meth}
In this section we will give a brief summary of our parallel implementation of the ES-Capon beamformer as presented in \cite{asen2012} and \cite{Buskenes}. This implementation was our starting point when the BS-Capon beamformer was implemented.

If the beamformer in (\ref{eq:z_mv}) is analyzed in combination with the matrix equation in (\ref{eq:b}) and the covariance estimation process in (\ref{eq:diag}), we see that the calculation of a set of Capon weights is independent across the outputs $z[n]$. As shown in Fig.\ \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently. Further, based on memory dependencies and to hold memory consumption per thread at a minimum, a group of threads should process one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes. That is, for each data vector $\vec{x}$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}.\\
\text{2) } &\text{Solve the linear system of equations as in (\ref{eq:b}}).\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}.
%\end{array}
\end{align*}
%\begin{enumerate}
%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
%\end{enumerate}
%\begin{enumerate}
%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
%\end{enumerate}
%\begin{enumerate}
%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
%\end{enumerate}
The computational complexities of these three steps are $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$, respectively. As $M$ (and thereby $L$) increases, the first two steps quickly become unmanageable, and real-time performance in cardiac ultrasound imaging can only be achieved by exploring methods with reduced complexity.

@@

In \cite{asen2012, Buskenes} we demonstrated how the cubic complexity with $L$ %and linear complexity with $N_K$ 
can be reduced by realizing that calculations in (\ref{eq:R}) overlap across subarrays. %Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ in a sliding window maner as
%\begin{align}
%\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
%&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
%&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
%\end{align}
%Along the time dimension we also find that
%\begin{align}
%\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
%&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
%\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
%Sliding windows across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$. However, since $N_K$ typically is small, and the GPU has a limited amount of shared memory available per thread, we focus on subarrays only. 
%In that case, 
The complexity of calculating $\mat{\hat{R}}$ is by this reduced to $O(N_K(L(N_L-2)+2L^2))$. For $L=M/2$ this reduces to $O(N_KM^2)$.
%The sliding-window approach will cause dependence among the iterative calculations. The level of task granularity is therefore increased. Calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$. One thread is therefore assigned to each diagonal in the upper triangle of $\mat{\hat{R}}$. For small values of $L$, several covariance matrices are calculated per thread block (Fig. \ref{fig:gpulayout}). 
Maximum supported $L$ is also increased to the maximum number of threads per block (currently 1024). %However, for small-$L$-high-$M$ configurations, memory consumption per thread will be too high to achieve full GPU occupancy.

%Taking symmetry into account, only the upper half of $\mat{\hat{R}}$ needs to be calculated. However, the GPU computes $\mat{\hat{R}}$ in such a way that the lower triangle can be found with little overhead \cite{Buskenes}. This also means that no special solver is needed in order to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if a symmetric solver could have saved some instructions.

%For the adaptive diagonal loading factor in (\ref{eq:diag}), the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory. One row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements therefore need to reside in shared memory until the trace is accumulated. When that is done, scattered writes of the scaled diagonal elements to global memory has to be performed.

@@

\subsection{Solving Multiple Small Linear Systems}
%The literature has mostly focused on large systems when solving systems of linear equations on the GPU. GPU libraries therefore often lack a comprehensive collection of batched solvers for large sets of small systems. The reason for this is two-fold. First the GPU needs to solve thousands of small systems in order to beat the low memory latency of the CPU. Second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of shared memory and registers are available. Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared to the CPU in every case.

In \cite{asen2012} we discussed how it is not given that solving on the GPU is faster than solving on the CPU in general. However, there are two arguments for solving on the GPU. First the CPU, which might already run a lot of ultrasound-related algorithms, is offloaded. Second, the transfer of several thousand covariance matrices over the PCI-Express bus, from the CPU to the GPU, is avoided.

@@

\subsection{Compute Beamformer Output}
%The calculation of the beamformer output in (\ref{eq:z_mv}) has been implemented using one thread per element weight ($L$ threads per weight vector). After the solution vector $\vec{b}$ has been placed in shared memory, one of the $L$ threads finds the sum of $\vec{b}$. 
%Another option is to let several threads cooperate to form the sum using a binary reduction sum. However, in the following benchmarks the first approach has been used. 
%Each of the $L$ threads then reads one element of $\vec{b}$, and scales it with the inverse sum to get the weight vector $\vec{w}[n]$. 
The final output can be formed by either reducing the data down to length $L$ as in (\ref{eq:z_mv}), or increasing the weight vector to length $M$ as in (\ref{eq:z_mv2}). Since we selected to use $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate 
\begin{align}\label{eq:subsum}
\breve{x}_{r} = \sum_{k=r}^{r+N_L-1}x_{k} \text{ for } r \in [0, L-1].
\end{align}
Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \frac{1}{N_L}\sum_{r=0}^{L-1} z_r$ to global memory. %The last sum can also be performed using a binary reduction sum across several threads for increased speed.

@@

The other option is to transform the element space covariance matrix to beamspace:
\begin{align}
\mat{\hat{R}}_\text{BS} &= \mat{B}\mat{\hat{R}}\mat{B}^H.
\end{align}
The advantage of this approach is that minor changes have to be made to the pipeline described in Section \ref{sec:meth}, but it will only speed up the solving step if the solution $\vec{b}_{BS}$ is transformed back to element space. In this paper we will focus on the first approach, and how it has been implemented on the GPU.
%By this approach we go from solving $\mat{\hat{R}}\vec{b} = \vec{1}_L$, which is an $L \times L$ system, to solving the $N_b \times N_b$ system $\mat{\hat{R}}_{BS}\vec{b}_{BS} = (\sqrt{L}\vec{e}_1)$. Now, one can either choose to transform the subarray sums, as in (\ref{eq:subsum}), to beamspace or transform the result, $\vec{b}_{BS}$, back to element space. 
In both cases solving has to be performed against a beamspace steering vector
\begin{align}
\vec{a}_\text{BS} &= \mat{B}\vec{1}_L = \sqrt{L}\vec{e}_1,
\end{align}
where $\vec{e}_i$ has the value one in the $i$th position and zeroes elsewhere. Note that the factor $\sqrt{L}$ was not included in \cite{Nilsen2009}.

As mentioned, transforming subarrays into beamspace yields subarrays that no longer overlap. This means that the spatial overlap previously exploited when calculating the sample covariance matrix in element space is no longer available. On the other hand, since $N_b$ should be small ($N_b < 32$), one thread assigned to each element in $\mat{\hat{R}}_\text{BS}$ is now an option. A kernel similar to the one described in \cite{Chen2011} has therefore been implemented. The computational complexity of this step is now $O(N_KN_LN_b^2)$, which for $L=M/2$ is less than the complexity of building $\mat{\hat{R}}$ in element space as long as $N_b < \sqrt{M}$.

%The implementation calculating covariance matrices therefore needs to be adjusted to handle non-overlapping subarrays. The complexity of constructing the covariance matrix is changed to $O(N_b^2L)$, so constructing the covariance matrices will be faster in beamspace as long as $N_b < \sqrt{L}$.
%Derivation of sliding beamspace transformation. 
%We will now show how to minimize the amount of operations required for the beamspace transformation by utilizing the overlap between subarrays previously exploited when estimating the covariance matrix in element space. 
The overlap between subarrays that were exploited when estimating the covariance matrix in element space, can now be utilized when calculating the beamspace transform. 
Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l$th subarray for the $k$th beam is from (\ref{eq:butler}) found to be:
\begin{align}\label{eq:bs_formula}
x_{\text{BS},l,k} &= \mat{B}_k\vec{x}_l \,\,\,\,\,\,\,\,\, k \in [0, N_b-1] \\
&= \frac{1}{\sqrt{L}}\sum_{p=i}^{l+L-1}x_p e^{-j2\pi k(p-l)/L},
\end{align}
where $\mat{B}_k$ is the $k$th row of $\mat{B}$. %Manipulating this expression we find that for $l \in [1, L-1]$
%\begin{align}\label{eq:sliding_bs}
%x_{\text{BS},l,k} &= \frac{1}{\sqrt{L}}((x_{\text{BS},l-1,k} - x_{l-1})e^{-j2\pi k/L} \nonumber \\ &+ x_{i+L-1}e^{-j2\pi k(L-1)/L}) \nonumber \\
%&= \frac{1}{\sqrt{L}}(x_{\text{BS},l-1,k} - x_{l-1} + x_{i+L-1})e^{-j2\pi k/L}.
%\end{align}

%We denote this the sliding-window beamspace transformation (SBS), and it is equal to a normalized sliding DFT \cite{Lyons2003}. For $l=0$ a full computation has to be performed using (\ref{eq:bs_formula}).
This formula has been implemented using a normalized sliding DFT (SDFT)\cite{Lyons2003}.
%Equation (\ref{eq:sliding_bs}) shows that 
The SDFT allows us to still save instructions by utilizing the overlap between subarrays, but with an increase in task level granularity. In this situation, it might be tempting to use one thread per beam (index $k$) that calculates this beam for all subarrays. However, since the number of selected beams usually is small ($N_b \ll M$) we will end up with a large memory-per-thread ratio ($M/N_b$). Using one thread per beam is therefore not feasible for small-$N_b$-large-$M$ configurations. One solution, as depicted in Fig.\ \ref{fig:gpulayoutbs}, is to combine SDFT with full computations. A number of sub thread groups of size $N_b$ is created that targets different sets of subarrays which they transform using the SDFT algorithm. In that way, the number of threads per sample is increased and thus the memory-per-thread ratio is decreased.

After solving $\vec{b}_\text{BS} = \sqrt{L}\mat{\hat{R}}_\text{BS}^{-1}\vec{e}_1$, the beamspace weight vector is calculated as 
\begin{align}
\vec{w}_\text{BS} = \frac{\vec{b}_\text{BS}}{\sqrt{L}\vec{e}_1\vec{b}_\text{BS}} \in \mathbb{C}^{N_b}
\end{align}  
using one thread per beam, and applied to the beamspace data summed across subarrays,
\begin{align}
z[n] = \vec{w}_\text{BS}^H\sum_{l=0}^{L-1}\vec{x}_{\text{BS},l}. 
\end{align}
%Several pixels are calculated per thread block for small values of $N_b$ in order to achieve high GPU occupancy. The beamspace data summed across subarrays are calculated as part of the SBS kernel.

>>>>

For a three dimensional beamspace we show that processing rates supporting real-time cardiac ultrasound imaging are possible, meaning that images can be processed faster than the image acquisition rate for a wide range of parameters. Image quality is investigated in videos of both a simulated phantom, and in an \textit{in-vivo} cardiac dataset. These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with improved lateral resolution and no local shift-variance effects, both in element-space and beamspace.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway.

@@

% denne må utelates i IUS-artikkelen
In recent work \cite{asen2012, Buskenes, Buskenes2013} we have applied a GPU-based Capon beamformer for both active sonar imaging and for cardiac ultrasound imaging. Sonar has a less strict real-time requirement than medical ultrasound imaging, which means that most array configurations can be handled in real time using the GPU. Capon beamforming in medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates that not even a high-end GPU can provide. Since the execution time is cubic with respect to the number of array elements, we need methods that reduce the computational complexity for large arrays. Various approaches to low-complexity Capon beamforming for medical ultrasound imaging have therefore been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}. In this paper we will make use of a beamspace transformation described by Nilsen and Hafizovic \cite{Nilsen2009}, in order to maintain high framerates as the number of elements increases. 
%The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging. 
%(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)

In the next section we present background information on Capon beamforming in both element-space and beamspace and the GPU computing model. In Section \ref{sec:meth} we recap our parallel implementation of the Capon beamformer, before the GPU implementation of the beamspace Capon beamformer is presented in Section \ref{sec:bs}. Benchmark results and resulting images are presented in Section \ref{sec:bench} and \ref{sec:images}, together with a discussion of trading image quality for speed in Section \ref{sec:trade}.  Finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.

@@

The conventional form of array beamforming in medical ultrasound imaging is delay-and-sum (DAS). Here, the signal recorded at element number $m \in [0,M-1]$ at time $n$ is delayed with an appropriate delay $\Delta_m[n]$ to focus and steer the ultrasound beam. This delayed data is here denoted $x_m[n]$. The beamformer output $z[n]$ is then the sum of all $M$ elements weighted with a factor $w_m$,
\begin{align}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n] = \vec{w}^H\vec{x}[n], \label{eq:z}
\end{align}
where $(\cdot)^H$ is the Hermitian transpose. The weight vector, or window, $\vec{w}$ is usually real in DAS beamforming, and is applied to trade resolution for a better signal-to-noise ratio (SNR). Although different windows can be used in different areas of the images, $\vec{w}$ is usually predefined.

As shown in Fig. \ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.   The Capon beamformer produces weights that minimize the beamformer output power while maintaining unity gain in the focus direction. The result is that interference impinging from other directions is suppressed \cite{Synnevag2007}. The complex weights cause a micro-steering which is minute compared to the initial steering and focusing step, therefore the beamformer also works well for broadband ultrasound signals.

@@

\begin{figure}
\centerline{\includegraphics[width=3in]{gfx/beamforming_mv.eps}}
\caption{Capon beamforming. Impinging signal is demodulated, and aligned in phase by applying steering and focusing delays $\Delta_m[n]$. Adaptive weights, $\vec{w}[n]$, are then calculated based on in-phase data, and finally the output is formed by a Capon-weighted coherent sum.}
\label{fig:mvbf}
\end{figure}

@@

For a focused system like cardiac ultrasound imaging, where the received signal is concentrated in a narrow sector around the transmit direction, a small percentage of the beams contains almost all the received energy. Nilsen and Hafizovic concluded that as little as three beams around (and including) the transmit direction were adequate to produce results comparable to applying Capon beamforming in element space. Hence, for equally pitched arrays the inversion steps is found to be constant with respect to the number of array elements.

@@

In CUDA programming a kernel function is executed over a large set of threads, where the kernel function describes the work to be done in each thread. As depicted in Fig. \ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block (e.g. $32\times32$). Threads inside a block support low-cost synchronization if needed. From a CUDA perspective the GPU consist of one or more streaming multi-processors (SM), where each SM is capable of 32 or more concurrent multiplications or additions, or both, depending on the architecture. A group of blocks is scheduled to each SM until all blocks have been processed.

@@

\section{Parallel ES-Capon}\label{sec:meth}
In this section we will recap our parallel implementation of the ES-Capon beamformer as presented in \cite{asen2012, Buskenes}. This implementation was our starting point when the BS-Capon beamformer was implemented.

If the beamformer in (\ref{eq:z_mv}) is analyzed in combination with the matrix equation in (\ref{eq:b}) and the covariance estimation process in (\ref{eq:diag}), we see that the calculation of a set of Capon weights is independent across the outputs $z[n]$. As shown in Fig. \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently. Further, based on memory dependencies and to hold memory consumption per thread at a minimum, a group of threads should process one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes. That is, for each data vector $\vec{x}$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}.\\
\text{2) } &\text{Solve the linear system of equations as in (\ref{eq:b}}).\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}.
%\end{array}
\end{align*}
%\begin{enumerate}
%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
%\end{enumerate}
%\begin{enumerate}
%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
%\end{enumerate}
%\begin{enumerate}
%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
%\end{enumerate}
The computational complexities of these three steps are $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$, respectively. As $M$ (and thereby $L$) increases, the first two steps quickly become unmanageable, and real-time performance in cardiac ultrasound imaging can only be achieved by exploring methods with reduced complexity.

@@

In \cite{asen2012, Buskenes} we demonstrated how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap across subarrays. Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ in a sliding window maner as
\begin{align}
\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
\end{align}
%Along the time dimension we also find that
%\begin{align}
%\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
%&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
%\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
Sliding windows across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$. However, since $N_K$ typically is small, and the GPU has a limited amount of shared memory available per thread, we focus on subarrays only. In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_K(L(N_L-2)+2L^2))$. For $L=M/2$ this reduces to $O(N_KM^2)$.

The sliding-window approach will cause dependence among the iterative calculations. The level of task granularity is therefore increased. Calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$. One thread is therefore assigned to each diagonal in the upper triangle of $\mat{\hat{R}}$. For small values of $L$, several covariance matrices are calculated per thread block (Fig. \ref{fig:gpulayout}). Maximum supported $L$ is now increased to the maximum number of threads per block (currently 1024). However, for small-$L$-high-$M$ configurations, memory consumption per thread will be too high to achieve full GPU occupancy.

Taking symmetry into account, only the upper half of $\mat{\hat{R}}$ needs to be calculated. However, the GPU computes $\mat{\hat{R}}$ in such a way that the lower triangle can be found with little overhead \cite{Buskenes}. This also means that no special solver is needed in order to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if a symmetric solver could have saved some instructions.

For the adaptive diagonal loading factor in (\ref{eq:diag}), the trace of $\mat{\hat{R}}$ has to be known before diagonal elements are written back to global memory. Since one row for all covariance matrices is written to global memory per kernel iteration, the diagonal elements need to be held in shared memory until the trace is accumulated. When that is done, scattered writes of the scaled diagonal elements to global memory has to be performed.

@@

\subsection{Solving Multiple Small Linear Systems}
The literature has mostly focused on large systems when solving systems of linear equations on the GPU. GPU libraries therefore often lack a comprehensive collection of batched solvers for large sets of small systems. The reason for this is two-fold. First the GPU needs to solve thousands of small systems in order to beat the low memory latency of the CPU. Second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of shared memory and registers are available. Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared to the CPU in every case. It is therefore not given that solving on the GPU is faster than solving on the CPU in general. However, there are two arguments for solving on the GPU. First the CPU, which might already run a lot of ultrasound-related algorithms, is offloaded. Second, the transfer of several thousand covariance matrices over the PCI-Express bus, from the CPU to the GPU, is avoided.

@@

\subsection{Compute Beamformer Output}
The calculation of the beamformer output in (\ref{eq:z_mv}) has been implemented using one thread per element weight ($L$ threads per weight vector). After the solution vector $\vec{b}$ has been placed in shared memory, one of the $L$ threads finds the sum of $\vec{b}$. 
%Another option is to let several threads cooperate to form the sum using a binary reduction sum. However, in the following benchmarks the first approach has been used. 
Each of the $L$ threads then reads one element of $\vec{b}$, and scales it with the inverse sum to get the weight vector $\vec{w}[n]$.

The final output can be formed by either reducing the data down to length $L$ as in (\ref{eq:z_mv}), or increasing the weight vector to length $M$ as in (\ref{eq:z_mv2}). Since the design is $L$ threads per output, the data is reduced to length $L$ by letting thread $r$ calculate 
\begin{align}\label{eq:subsum}
\breve{x}_{r} = \sum_{k=r}^{r+N_L-1}x_{k} \text{ for } r \in [0, L-1].
\end{align}
Each thread now computes $z_r = w_r^*x_r$, and finally one thread computes and outputs $z = \frac{1}{N_L}\sum_{r=0}^{L-1} z_r$ to global memory. %The last sum can also be performed using a binary reduction sum across several threads for increased speed.

@@

The other option is to transform the element space covariance matrix to beamspace:
\begin{align}
\mat{\hat{R}}_\text{BS} &= \mat{B}\mat{\hat{R}}\mat{B}^H.
\end{align}
The advantage of this approach is that minor changes have to be made to the pipeline described in Section \ref{sec:meth}, but it will only speed up the solving step if the solution $\vec{b}_{BS}$ is transformed back to element space. In this paper we will focus on the first approach, and how it has been implemented on the GPU.
%By this approach we go from solving $\mat{\hat{R}}\vec{b} = \vec{1}_L$, which is an $L \times L$ system, to solving the $N_b \times N_b$ system $\mat{\hat{R}}_{BS}\vec{b}_{BS} = (\sqrt{L}\vec{e}_1)$. Now, one can either choose to transform the subarray sums, as in (\ref{eq:subsum}), to beamspace or transform the result, $\vec{b}_{BS}$, back to element space. 
In both cases solving has to be performed against a beamspace steering vector
\begin{align}
\vec{a}_\text{BS} &= \mat{B}\vec{1}_L = \sqrt{L}\vec{e}_1,
\end{align}
where $\vec{e}_i$ has the value one in the $i$th position and zeroes elsewhere.

As mentioned, transforming subarrays into beamspace yields subarrays that no longer overlap. This means that the spatial overlap previously exploited when calculating the sample covariance matrix in element space is no longer available. On the other hand, since $N_b$ should be small ($N_b < 32$), one thread assigned to each element in $\mat{\hat{R}}_\text{BS}$ is now an option. A kernel similar to the one described in \cite{Chen2011} has therefore been implemented. In addition, for small $N_b$ values, several covariance matrices are calculated per thread block to achieve high GPU occupancy. The computational complexity of this step is now $O(N_KN_LN_b^2)$, which for $L=M/2$ is less than the complexity of building $\mat{\hat{R}}$ in element space as long as $N_b < \sqrt{M}$.

%The implementation calculating covariance matrices therefore needs to be adjusted to handle non-overlapping subarrays. The complexity of constructing the covariance matrix is changed to $O(N_b^2L)$, so constructing the covariance matrices will be faster in beamspace as long as $N_b < \sqrt{L}$.
%Derivation of sliding beamspace transformation. 
We will now show how to minimize the amount of operations required for the beamspace transformation by utilizing the overlap between subarrays previously exploited when estimating the covariance matrix in element space. Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l$th subarray for the $k$th beam is from (\ref{eq:butler}) found to be:
\begin{align}\label{eq:bs_formula}
x_{\text{BS},l,k} &= \mat{B}_k\vec{x}_l \,\,\,\,\,\,\,\,\, k \in [0, N_b-1] \\
&= \frac{1}{\sqrt{L}}\sum_{p=i}^{l+L-1}x_p e^{-j2\pi k(p-l)/L},
\end{align}
where $\mat{B}_k$ is the $k$th row of $\mat{B}$. Manipulating this expression we find that for $l \in [1, L-1]$
\begin{align}\label{eq:sliding_bs}
x_{\text{BS},l,k} &= \frac{1}{\sqrt{L}}((x_{\text{BS},l-1,k} - x_{l-1})e^{-j2\pi k/L} \nonumber \\ &+ x_{i+L-1}e^{-j2\pi k(L-1)/L}) \nonumber \\
&= \frac{1}{\sqrt{L}}(x_{\text{BS},l-1,k} - x_{l-1} + x_{i+L-1})e^{-j2\pi k/L}.
\end{align}
We denote this the sliding-window beamspace transformation (SBS), and it is equal to a normalized sliding DFT \cite{Lyons2003}. For $l=0$ a full computation has to be performed using (\ref{eq:bs_formula}).

Equation (\ref{eq:sliding_bs}) shows that instructions are saved if we utilize the overlap between subarrays, but with an increase in task level granularity (as discussed in Section \ref{sec:calcR}). In this situation, it might be tempting to use one thread per beam (index $k$) that calculates this beam for all subarrays. However, since the number of selected beams usually is small ($N_b \ll M$) we will end up with a large memory-per-thread ratio ($M/N_b$). Using one thread per beam is therefore not feasible for small-$N_b$-large-$M$ configurations. One solution, as depicted in Fig. \ref{fig:gpulayoutbs}, is to combine SBS with full computations. A number of sub thread groups of size $N_b$ is created that targets different sets of subarrays which they transform using the SBS algorithm. In that way, the number of threads per sample is increased and thus the memory-per-thread ratio is decreased.

After solving $\vec{b}_\text{BS} = \sqrt{L}\mat{\hat{R}}_\text{BS}^{-1}\vec{e}_1$, the beamspace weight vector is calculated as 
\begin{align}
\vec{w}_\text{BS} = \frac{\vec{b}_\text{BS}}{\sqrt{L}\vec{e}_1\vec{b}_\text{BS}} \in \mathbb{C}^{N_b}
\end{align}  
using one thread per beam, and applied to the beamspace data summed across subarrays,
\begin{align}
z[n] = \vec{w}_\text{BS}^H\sum_{l=0}^{L-1}\vec{x}_{\text{BS},l}. 
\end{align}
Several pixels are calculated per thread block for small values of $N_b$ in order to achieve high GPU occupancy. The beamspace data summed across subarrays are calculated as part of the SBS kernel.
